Recalibrating Gravitational Wave Phenomenological Waveform Model

### 2023-05-26 src/tex/ms.tex
update

With the loss function evaluated, we apply gradient descent to optimize the tunable
coefficients as shown in Algorithm~\ref{alg:gradient}. We take $\lambda_i$ to be the 
original coefficients given in \citep{khan2016frequency}. We take them as the initial 
waveform coefficients because they lie in the neighborhood of the minimum that we wish 
to find. Then, by taking $\alpha=10^{-6}$, we see in Fig. \ref{fig:loss} plateau at 
around $N=30000$, hence we end the optimization here.  
%\te{I think we need a bit more discussion about the stopping criterion here. What is N, and why did we choose it to be this number? Ideally we would actually have a plot of the loss function during training. I don't have a good intuition for if its noisy. Does it plateau? If we wanted to ensure that people don't complain about overfitting we could also plot the loss of the validation set to show its not going up.}

@@

To evaluate the effectiveness of the optimization, we analyze approximately 530
NR waveforms \kw{Just say how many you have} from the SXS catalog in this study.
We specifically select waveforms with negligible eccentricity
(${e<2\times10^{-3}}$) and precession (${ \chi_{x,y}<5\times10^{-3}}$) that are
consistent with the constraints of the waveform model. As shown in
Fig.~\ref{fig:intrin_space}, the intrinsic parameters of the chosen test
waveforms fall within the parameter space defined by the training waveforms.
This suggests that these test waveforms can be used for direct comparison with
the original model.

@@

\begin{figure}[t]
	\script{all_quadrants.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/all_quadrants.pdf}
	\caption{Distributions of mismatches in the top-left and bottom-right regions.
	We use a constant noise spectrum to calculate mismatch and
	$\mathcal{L}_{\mathrm{mean}}$ as the loss function. Waveforms in the top-left region generally improves while waveforms in the bottom-right region worsened, as indicated by the median (dashed lines).}
	\label{fig:all_quadrants}
\end{figure}

\begin{figure}[t]
	\script{ps_q148_quadrants.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/ps_q148_quadrants.pdf}
	\caption{Parameter space of testing waveforms. Each region is fitted independently with waveforms listed in Tab.~\ref{tab:quadrants}. Colorbar represents the $\log_{10}$ difference of mismatches before and after optimization.}
	\label{fig:ps_q148_quadrant}
\end{figure}

@@

The results presented in Fig.~\ref{fig:q148_q1248_compare} demonstrate that
increasing the number of training waveforms used in waveform optimization yields
only a marginal increase in accuracy. Our analysis suggests that this marginal
improvement is a consequence of over-determination of the waveform coefficients.
Consequently, increasing the number of calibration NR waveforms is unlikely to
result in any significant improvement of the model's accuracy. These
observations suggest that the parameterized ansatz employed in our study may not
be suitable for certain regions in the parameter space, leading to mismatches
for some waveforms while other waveforms remain at the high mismatch tail with
negligible changes. This highlights the constraints of the model's flexibility
that ultimately limit its performance.

The reduced spin approximation is a major contributor to the inaccuracies
observed in the ansatz. In IMRPhenomD, this approximation employs a single spin
parameter, $\chi_{\mathrm{PN}}$, as described in Sec. \ref{sec:method}. The
parameterization of BBH mergers using a single spin parameter results in a
degeneracy within the parameter space. Specifically, black hole events with
different spins may generate the same waveform due to identical values of
$\chi_{\mathrm{PN}}$, leading to erroneous results, particularly for highly
unequal spin events. This degeneracy produces straight lines in the parameter
space with negative slopes that depend on the mass ratio, which can be seen in 
Fig.\ref{fig:ps_q148}. Notably, the ansatz performs better in the top-left
region than the bottom-right region. In an attempt to address
this issue, we partitioned the parameter space into four regions, as described
in Sec. \ref{sec:result}. With separate optimizations for each
regions, Fig.\ref{fig:ps_q148_quadrant} indicates that the top-left region's 
performance has improved, while the bottom-right region hardly improves. 
This observation suggests the ansatz is specific to certain regions of the 
parameter space, with a preference for BBH events lying above the diagonal, 
and it has limited enhancement for events lying below the diagonal.

>>>>

With the loss function evaluated, we apply gradient descent to optimize the tunable
coefficients as shown in Algorithm~\ref{alg:gradient}. We take $\lambda_i$ to be the 
original coefficients given in \citep{khan2016frequency}. We take them as the initial 
waveform coefficients because they lie in the neighborhood of the minimum that we wish 
to find. Then, by taking $\alpha=10^{-6}$, we see the validation loss in 
Fig.~\ref{fig:loss} plateau at around $N=12000$, hence we end the optimization here.  
%\te{I think we need a bit more discussion about the stopping criterion here. What is N, and why did we choose it to be this number? Ideally we would actually have a plot of the loss function during training. I don't have a good intuition for if its noisy. Does it plateau? If we wanted to ensure that people don't complain about overfitting we could also plot the loss of the validation set to show its not going up.}

@@

To evaluate the effectiveness of the optimization, we analyze 536 NR waveforms 
from the SXS catalog in this study. We specifically select waveforms with 
negligible eccentricity (${e<2\times10^{-3}}$) and precession 
(${\chi_{x,y}<5\times10^{-3}}$) that are consistent with the constraints of 
the waveform model. As shown in Fig.~\ref{fig:intrin_space}, the intrinsic 
parameters of the chosen test waveforms fall within the parameter space defined 
by the training waveforms. This suggests that these test waveforms can be used 
for direct comparison with the original model.

@@

\begin{figure}[t]
	\script{all_quadrants.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/all_quadrants.pdf}
	\caption{Distributions of mismatches in the top-left (Top) and bottom-right (Bottom) regions.
	We use a constant noise spectrum to calculate mismatch and
	$\mathcal{L}_{\mathrm{mean}}$ as the loss function. Waveforms in the top-left region generally improves while waveforms in the bottom-right region has very little improvement, as indicated by the median (dashed lines).}
	\label{fig:all_quadrants}
\end{figure}

\begin{figure}[t]
	\script{ps_q148_quadrants.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/ps_q148_quadrants.pdf}
	\caption{Parameter space of testing waveforms. Each region is fitted independently with waveforms listed in Tab.~\ref{tab:quadrants}.}
	\label{fig:ps_q148_quadrant}
\end{figure}

@@

The results presented in Fig.~\ref{fig:q148_q1248_compare} demonstrate that
increasing the number of training waveforms used in waveform optimization yields
only a marginal increase in accuracy. Our analysis suggests that this marginal
improvement is a consequence of over-determination of the waveform coefficients.
Consequently, increasing the number of calibration NR waveforms is unlikely to
result in any significant improvement of the model's accuracy. These
observations suggest that the parameterized ansatz employed in our study may not
be suitable for certain regions in the parameter space, leading to low mismatches
for some waveforms while other waveforms remain at the high mismatch tail with
negligible changes. This highlights the constraints of the model's flexibility
that ultimately limit its performance.

The reduced spin approximation is a major contributor to the inaccuracies
observed in the ansatz. In IMRPhenomD, this approximation employs a single spin
parameter, $\chi_{\mathrm{PN}}$, as described in Sec.~\ref{sec:method}. The
parameterization of BBH mergers using a single spin parameter results in a
degeneracy within the parameter space. Specifically, black hole events with
different spins may generate the same waveform due to identical values of
$\chi_{\mathrm{PN}}$, leading to erroneous results, particularly for highly
unequal spin events. This degeneracy produces straight lines in the parameter
space with negative slopes that depend on the mass ratio, which can be seen in 
Fig.~\ref{fig:ps_q148}. Notably, the ansatz performs better in the top-left
region than the bottom-right region. In an attempt to address
this issue, we partitioned the parameter space into four regions, as described
in Sec.~\ref{sec:result}. With separate optimizations for each
regions, Fig.~\ref{fig:ps_q148_quadrant} indicates that the top-left region's 
performance has improved, while the bottom-right region hardly improves. 
This observation suggests the ansatz is specific to certain regions of the 
parameter space, with a preference for BBH events lying above the diagonal, 
and it has limited enhancement for events lying below the diagonal.
